Fractional Types

### 2016-12-15 typespp.tex
up to end sec 3

\begin{definition}[Power of a 1-combinator]
  The $k^{th}$ power of a 1-combinator $p : \tau \iso \tau$, for
  $k \in \Z$ is
  \[
    p^k =
  \begin{cases}
    \idiso & k = 0 \\
    p \odot p^{k - 1} & k > 0 \\
    (!~p) \odot p^{k + 1} & k < 0 \\
  \end{cases}
  \]
\end{definition}

@@

From a denotational perspective, a $\Pi$ type $\tau$ denotes a finite set, a
$\Pi$ 1-combinator denotes a permutation on finite sets, and the 2-combinators
denote coherence conditions on these permutations~\cite{Carette2016}. Formally,
the language $\Pi$ is a \emph{categorification}~\cite{math/9802029} of the
natural numbers as a \emph{symmetric rig groupoid}~\cite{nlabrig}. This
structure is a \emph{symmetric bimonoidal category} or a \emph{commutative rig
category} in which every morphism is invertible. The underlying category
consists of two symmetric monoidal structures~\cite{nla.cat-vn1051288}
separately induced by the properties of addition and multiplication of the
natural numbers. The monoidal structures are then augmented with distributivity
and absorption natural isomorphisms~\cite{laplaza} to model the full commutative
semiring (aka, commutative rig) of the natural numbers. Despite this rich
structure, the individual objects in the category for $\Pi$ are just plain sets
with no interesting structure. In this section we introduce, in the denotation
of $\Pi$, some non-trivial groupoids which we call \emph{iteration groupoids}
and \emph{symmetry groupoids}. Under certain conditions, sums and products of
these groupoids behave as expected which ensures that a sensible compositional
programming language can be designed around them.

@@

The previous construction is quite useful: it allows us to take a set of
cardinality $N$ and a permutation on that set of order $P$ to construct a
groupoid of cardinality $\frac{N}{P}$. Although this idea allows us to construct
a groupoid of cardinality $\frac{3}{2}$ as shown above, it is not expressive
enough to construct a groupoid of cardinality $\frac{1}{3}$. Indeed if the
underlying set has only one element, the only permutation is the identity and
$P$ must be 1. The construction, however, already contains the main ingredient
needed for the construction of more general groupoids with fractional
cardinality. This key piece is the set of iterates of a combinator which we will
use in two different ways: (i) a perspective of combinators-as-data in which the
iterates of a combinator $p$ become the objects of groupoid $\order{p}$ of
cardinality $\ord{p}$, and (ii) a perspective of combinators-as-symmetries in
which the iterates of a combinator $p$ become the internal symmetries of a
one-element groupoid $\iorder{p}$ of cardinality $\frac{1}{\ord{p}}$. We explain
these two perspectives in this section and the next.

We define the $k^{\text{th}}$ iterate of $p$ as:

and then collect all such iterates into a type

\noindent For example, the zeroth and first iterate of a combinator are represented as

These ingredients are sufficient to construct a groupoid $\order{p}$ from the
iterates of a 1-combinator $p : \tau\iso\tau$. The objects will be
the triples $\triple{k}{q}{\alpha}$ indexed by integers $k$, 1-combinators $q :
\tau\iso\tau$, and 2-combinators $\alpha : q \isotwo p^k$. This triple encodes
our knowledge that we have some (arbitrary) iterate $q$ of $p$; we do not have
any a priori knowledge of the actual syntactic structure of $q$, but we do
know that it is equivalent to $p ^ k$.

We then add (reversible!) morphisms between any iterates related by
2-combinators; categorically, this will make any such objects equivalent.
If $p$ has order $o$, Lem.~\ref{lem:ordercancel} gives us a
2-combinator $\alpha$ which witnesses that $p^i \isotwo p^{i+o}$.
Thus given two iterates $\triple{i}{q_i}{\alpha_i}$ and
$\triple{i+o}{q_j}{\alpha_j}$, they must be equivalent since
$\alpha_i~\bullet~\alpha~\bullet~!\,\alpha_j$ shows that $q_i \isotwo q_j$.
In other words $p^j$ will be equivalent to $p^k$ exactly when $j$ and $k$
differ by $o$.  This informal description formalizes straightforwardly:

@@

  \path[ultra  thick] (A) edge [my loop] node[below] {\texttt{id}} (A);
  \path[ultra  thick] (B) edge [my loop] node[above] {\texttt{id}} (B);
  \path[ultra  thick] (C) edge [my loop] node[below] {\texttt{id}} (C);
  \path[ultra  thick] (D) edge [my loop] node[above] {\texttt{id}} (D);
  \path[ultra  thick] (E) edge [my loop] node[below] {\texttt{id}} (E);

@@

The elements of $\iter{p}$ form a group under the following operation:
\[
\triple{k_1}{p_1}{\alpha_1} ~\circ~ \triple{k_2}{p_2}{\alpha_2} =
 \triple{k_1+k_2}{p_1 \odot p_2}{(\alpha_1 ~\respstwo~
    \alpha_2)~\transtwo~(\distiterplus{p}{k_1}{k_2})}
\]
where $\distiterplus{p}{k_1}{k_2}$ is defined in
Lem.~\ref{lem:distiterplus}. The common categorical representation of
a group is a category with one trivial object and the group elements
as morphisms on that trivial object. Our construction is essentially
the same except that the trivial object is represented as the iterates
of the identity over a singleton type.

Formally, given a 1-combinator $p : \tau\iso\tau$, the groupoid
$\iorder{p}$ is the groupoid (2-groupoid technically) whose objects
are the iterates in $\iter{\idiso_{\top}}$. Every pair of objects is
connected by the morphisms in $\iter{p}$. To capture the relationship
between these morphisms, we also have 2-morphisms relating the
distinct iterates as in the previous section.

@@

@@

  \path[ultra  thick] (A) edge [my loop] node[below] {$\idiso$} (A);
  \path[ultra  thick] (A) edge [out=-140, in=-40, looseness=20] node[below] {$\permtwo$} (A);

  \path[ultra  thick] (A) edge [bend left=50] node[above] {$\alpha_{-2,0}$} (B);
  \path[ultra  thick] (B) edge [bend left=50] node[above] {$\alpha_{0,2}$} (C);
  \path[ultra  thick] (C) edge [bend left=50] node[above] {$\alpha_{0,2}$} (D);
  \path[ultra  thick] (D) edge [bend left=50] node[above] {$\alpha_{0,2}$} (E);
\end{tikzpicture}
\end{center}

%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Sums and Products of Iteration and Symmetry Groupoids}

If we are to build a compositional programming language around iteration and
symmetry groupoids, we need to ensure that they compose sensibly with the
existing type formers. Groupoids, viewed as categories, come associated with
natural notions of sums and products, and one might expect or hope that
identities which hold for rational numbers lift to identities in our
situation. The most prominent and important identity we seek is:
such as:
\[
\order{p} \otimes \iorder{p} \simeq \mathbb{1}
\]
The groupoids on either side are not categorically equivalent but they
do have the same cardinality. Thus if we focus on
cardinality-preserving morphisms, it is sensible to consider them
equivalent. The situation is subtle, however, and studied in detail in
the next section.

@@

>>>>

\begin{definition}[Iterated Power of a 1-combinator]
  The $k^{th}$ iterated power of a 1-combinator $p : \tau \iso \tau$, for
  $k \in \Z$ is
  \[
    p^k =
  \begin{cases}
    \idiso & k = 0 \\
    p \odot p^{k - 1} & k > 0 \\
    (!~p) \odot p^{k + 1} & k < 0 \\
  \end{cases}
  \]
\end{definition}

@@

From a denotational perspective, a $\Pi$ type $\tau$ denotes a finite set, a $\Pi$ 1-combinator denotes a permutation on finite sets, and the 2-combinators denote coherence conditions on these permutations~\cite{Carette2016}. Formally, the language $\Pi$ is a \emph{categorification}~\cite{math/9802029} of the natural numbers as a \emph{symmetric rig groupoid}~\cite{nlabrig}. This structure is a \emph{symmetric bimonoidal category} or a \emph{commutative rig category} in which every morphism is invertible. The underlying category consists of two symmetric monoidal structures~\cite{nla.cat-vn1051288} separately induced by the properties of addition and multiplication of the natural numbers. The monoidal structures are then augmented with distributivity and absorption natural isomorphisms~\cite{laplaza} to model the full commutative semiring (aka, commutative rig) of the natural numbers. Despite this rich structure, the individual objects in the category for $\Pi$ are just plain sets with no interesting structure. In this section we introduce, in the denotation of $\Pi$, some non-trivial groupoids which we call \emph{iteration groupoids} and \emph{symmetry groupoids}. Products of these groupoids behave as expected which ensures that a sensible compositional programming language can be designed around them.

@@

The previous construction is quite useful: it allows us to take a set of
cardinality $N$ and a permutation on that set of order $P$ to construct a
groupoid of cardinality $\frac{N}{P}$. Although this idea allows us to construct
a groupoid of cardinality $\frac{3}{2}$ as shown above, it is not expressive
enough to construct a groupoid of cardinality, say, $\frac{1}{3}$. Indeed if the
underlying set has only one element, the only permutation is the identity and
$P$ must be 1.

The construction, however, already contains the main ingredient needed for the construction of more general groupoids with fractional cardinality. This key piece is the set of iterates of a combinator which we formally define as follows.

\begin{definition}[$\iter{p}$]
For each 1-combinator $p : \tau\iso\tau$, we form the set $\iter{p}$ whose elements are triples consisting of an integer $k$, a 1-combinator $q : \tau\iso\tau$ and a 2-combinator $\alpha : q \isotwo p^k$.
\end{definition}
Each triple encodes our knowledge that we have some (arbitrary) iterate $q$ of $p$; we do not have any a priori knowledge of the actual syntactic structure of $q$, but we do know that it is equivalent to $p^k$. For example, we have:
\[\begin{array}{rcl}
 \iter{\permtwo} &=& \{ \triple{0}{\idiso}{\idisotwo},\triple{1}{\permtwo}{\idrdr}, \triple{-1}{\permtwo}{\idisotwo}, \ldots \}
\end{array}\]
The idea is that $\iter{\permtwo}$ is, up to equivalence, the set of all distinct iterates $(\permtwo)^k$ of $\permtwo$.  Because of the underlying group structure of automorphisms, there are, up to equivalence, only $\ord{\permtwo}$ distinct iterates in $\iter{\permtwo}$.  In a proof-irrelevant setting, $\iter{\permtwo}$ is simply $\{ \idiso, \permtwo \}$. In this section and the next, we will use the elements of $\iter{p}$ in two groupoid constructions as either objects (emphasizing their ``data'' aspect) or morphisms (emphasizing their ``symmetry'' aspect).

Given a 1-combinator $p : \tau\iso\tau$, we define the groupoid $\order{p}$ as follows. The objects are the elements of $\iter{p}$, i.e., the triples $\triple{k}{q}{\alpha}$ indexed by integers $k$, 1-combinators $q : \tau\iso\tau$, and 2-combinators $\alpha : q \isotwo p^k$. We then add (reversible) morphisms between any iterates related by 2-combinators; categorically, this will make any such objects equivalent.  If $p$ has order $o$, Lem.~\ref{lem:ordercancel} gives us a 2-combinator $\alpha$ which witnesses that $p^i \isotwo p^{i+o}$.  Thus given two iterates $\triple{i}{q_i}{\alpha_i}$ and $\triple{i+o}{q_j}{\alpha_j}$, they must be equivalent since $\alpha_i~\bullet~\alpha~\bullet~!\,\alpha_j$ shows that $q_i \isotwo q_j$.  In other words $p^j$ will be equivalent to $p^k$ exactly when $j$ and $k$ differ by $o$.  This informal description formalizes straightforwardly.

\begin{definition}[$\order{p}$] For each 1-combinator $p : \tau\iso\tau$, we form the groupoid $\order{p}$ as follows:
\begin{itemize}
\item objects are the elements $\triple{k}{q}{\alpha}$ of $\iter{p}$;
\item there is a morphism between $\triple{k_1}{q_1}{\alpha_1}$ and $\triple{k_2}{q_2}{\alpha_2}$ for each $\alpha : q_1 \isotwo q_2$
\end{itemize}
\end{definition}

@@

  \path[ultra  thick] (A) edge [my loop] node[below] {\texttt{id}} (A);
  \path[ultra  thick] (B) edge [my loop] node[below] {\texttt{id}} (B);
  \path[ultra  thick] (C) edge [my loop] node[below] {\texttt{id}} (C);
  \path[ultra  thick] (D) edge [my loop] node[below] {\texttt{id}} (D);
  \path[ultra  thick] (E) edge [my loop] node[below] {\texttt{id}} (E);

@@

The elements of $\iter{p}$ form a group under the following operation:
\[
\triple{k_1}{p_1}{\alpha_1} ~\circ~ \triple{k_2}{p_2}{\alpha_2} =
 \triple{k_1+k_2}{p_1 \odot p_2}{(\alpha_1 ~\respstwo~
    \alpha_2)~\transtwo~(\distiterplus{p}{k_1}{k_2})}
\]
where $\distiterplus{p}{k_1}{k_2}$ is defined in Lem.~\ref{lem:distiterplus}. The common categorical representation of a group is a category with one trivial object and the group elements as morphisms on that trivial object. Our construction of the groupoid $\iorder{p}$ is essentially the same.

\begin{definition}[$\iorder{p}$] For each 1-combinator $p : \tau\iso\tau$, we form the 2-groupoid $\iorder{p}$ as follows:
\begin{itemize}
\item the objects are the iterates of the identity combinator on $\tau$;
\item the morphisms between every pair of objects are the elements of $\iter{p}$;
\item there is a 2-morphisms between between 1-morphisms $\triple{k_1}{q_1}{\alpha_1}$ and $\triple{k_2}{q_2}{\alpha_2}$ for each $\alpha : q_1 \isotwo q_2$ for every $\alpha : q_1 \isotwo q_2$
\end{itemize}
\end{definition}

@@

As an example, the groupoid $\iorder{(\permtwo)}$ can be represented as follows.

\begin{center}
\begin{tikzpicture}[scale=0.6,every node/.style={scale=0.6}]
  \draw[dashed] (0,-0.3) ellipse (9cm and 3cm);

@@

  \path[ultra  thick] (A) edge [my loop] node[below] {$\idiso$} (A);
  \path[ultra  thick] (B) edge [my loop] node[below] {$\idiso$} (B);
  \path[ultra  thick] (C) edge [my loop] node[below] {$\idiso$} (C);
  \path[ultra  thick] (D) edge [my loop] node[below] {$\idiso$} (D);
  \path[ultra  thick] (E) edge [my loop] node[below] {$\idiso$} (E);

  \path[ultra  thick] (A) edge [bend left=50] node[above] {$\alpha_{-2,0}$} (B);
  \path[ultra  thick] (B) edge [bend left=50] node[above] {$\alpha_{0,2}$} (C);
  \path[ultra  thick] (C) edge [bend left=50] node[above] {$\alpha_{0,2}$} (D);
  \path[ultra  thick] (D) edge [bend left=50] node[above] {$\alpha_{0,2}$} (E);

  \path[ultra  thick] (A) edge [out=-140, in=-40, looseness=10] node[below] {$\permtwo$} (A);
  \path[ultra  thick] (B) edge [out=-140, in=-40, looseness=10] node[below] {$\permtwo$} (B);
  \path[ultra  thick] (C) edge [out=-140, in=-40, looseness=10] node[below] {$\permtwo$} (C);
  \path[ultra  thick] (D) edge [out=-140, in=-40, looseness=10] node[below] {$\permtwo$} (D);
  \path[ultra  thick] (E) edge [out=-140, in=-40, looseness=10] node[below] {$\permtwo$} (E);
\end{tikzpicture}
\end{center}

%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Products of Iteration and Symmetry Groupoids}

If we are to build a compositional programming language around iteration and
symmetry groupoids, we need to ensure that they compose sensibly with the
existing type formers. Groupoids, viewed as categories, come associated with
natural notions of sums and products, and one might expect or hope that
identities which hold for rational numbers lift to identities in our
situation. The most prominent and important identity we seek is:
such as:
\[
\order{p} \otimes \iorder{p} \simeq \mathbb{1}
\]
The groupoids on either side are not categorically equivalent but they do have the same cardinality. Thus if we focus on some notion of cardinality-preserving morphisms, it is sensible to consider them equivalent. The situation is however subtle and does not directly support a notion of sums. It is studied in detail in the next section.

@@

***WAVEFRONT***
